---
title: "Report On Malnutrition of Children Under the Age 5yrs in Kenya"
author: "Bore Laban,"
date: "5/10/2020"
output:
  html_document:
    number_sections: yes
  pdf_document: default
---

```{r message=FALSE, loaded-packages}
#load packages
library(tidyverse)
set.seed(12345)
library(here)
library(kableExtra)
library(ggplot2)
library(caret)
```

# **Introduction**
The report explores the malnutrition of children under the age of 5 years. The malnutrition data was collect from Kenya Demographic and Health Survey 2014 with 18901 observations and 4 variables of interest.

The goal of this report is to **develop an algorithm for assessing and predicting** the nutritional status of children in rural geographies of Kenya. We will answer the following three questions:

* Whether each of the dependent variables are normally distributed.
* Explore each of the dependent variables and determine the subset that best explains the variability in each of the response variable using stepwise regression method.
* Distinguish between the full model and the reduced model obtained during stepwise regression using mean error(ME), root mean square error(RMSE) and the mean absolute percentage error(MAPE)

# **Data Analysis**
Description of variables

## Outcome variables

```{r echo=F}
#select variables of interest
tibble(outcomeVar=c("bmiz06", "haz06", "waz06", "whz06"),
       labelVar=c("BMI-for-age Z-score","Length/height-for-age Z-score","Weight-for-age Z-score","Weight-for-length/height Z-score"),
       type=rep("continuous", rep=4)) %>% kable() %>% kable_styling(fixed_thead = T)
```

## Independent variables

```{r echo=F, comment=NA}
tibble(independentVar=c(
                         "b4",
                         "bidx",
                         "bord",
                         "v012",
                         "v024",
                         "v025",
                         "v106",
                         "v130",
                         "v131",
                         "v151",
                         "v152",
                         "v191"
                        ),
       LabelVar = c(
                         "sex of child",
                         "birth column number",
                         "birth order number",
                         "respondent's current age",
                         "region",
                         "type of place of residence",
                         "highest educational level",
                         "religion",
                         "ethnicity",
                         "sex of household head",
                         "age of household head",
                         "wealth index factor score (5 decimals)"
                         ),
       Type = c(
                          "Categorical",
                          "NA",
                          "Ordinal",
                          "Continuous",
                          "Categorical",
                          "Categorical",
                          "Categorical",
                          "Categorical",
                          "Categorical",
                          "Categorical",
                          "Continuous",
                          "Continuous"
                          )

) %>% kable() %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Data Preparation 

Selected 5 rows of dataset and all variables of interest  
```{r malnutrition-dataser, echo=FALSE}
malnutritionData <- read.csv("malnutritiondata.csv") %>% as_tibble()
#select variable of interest
malnutritionDataSet <- malnutritionData %>% select(-c(1:5), -v013, -b0) 
#First 5 dataset
kable(malnutritionDataSet[1:5,]) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", font_size = 7))

#select independent variables objects

#indepVar <- malnutritionDataSet %>% select(-c(13:16)) %>% view()

#select dependent variables objects

dep_haz06 <- malnutritionDataSet %>% select(haz06)  %>% unlist() %>% as.numeric()
dep_waz06 <- malnutritionDataSet %>% select(waz06)  %>% unlist() %>% as.numeric()
dep_whz06 <- malnutritionDataSet %>% select(whz06)  %>% unlist() %>% as.numeric()
dep_bmiz06 <- malnutritionDataSet %>% select(bmiz06) %>% unlist() %>% as.numeric()
```

## Explorative Data Analysis

**Summary of the dataset**
```{r,echo=F}
summary(malnutritionDataSet) %>% kable() %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", font_size = 5, fixed_thread=T, full_width = F))
```


### Checking for normality of each dependent variable

* Is each of the dependent variables normally distributed?

 **Using Histogram**
```{r hist_check_normality, echo=F}
# Is each of the dependent variables normally distributed
par(mfrow=c(2,2))
hist(dep_haz06, probability = T, main = "Length/height-for-age Z-score", xlab = "haz06")
  lines(density(dep_haz06))
hist(dep_waz06, probability = T,main = "Weight-for-age Z-score", xlab = "waz06")
  lines(density(dep_waz06))
hist(dep_whz06, probability = T,main = "Weight-for-length/height Z-score", xlab = "whz06")
  lines(density(dep_whz06))
hist(dep_bmiz06, probability = T, main =  "BMI-for-age Z-score", xlab = "bmiz06")
  lines(density(dep_bmiz06))
ggsave("img/hist_check_normality.png")
dev.off()
```

**Using Normal Probability Plot of Dependent Variables**

```{r, echo=F}
par(mfrow=c(2,2))
qqnorm(dep_haz06, main = "Length/height-for-age Z-score")
  qqline(dep_haz06)
qqnorm(dep_waz06, main =  "Weight-for-age Z-score")
  qqline(dep_waz06)
qqnorm(dep_whz06, main = "Weight-for-length/height Z-score")
  qqline(dep_whz06)
qqnorm(dep_bmiz06, main = "BMI-for-age Z-score")
  qqline(dep_bmiz06)  
```

### Variable selection and Stepwise Regression Method

* Explore each of the dependent variables and determine the subset that best explains the variability in each of the response variable using stepwise regression method 

**Variable Selection**


```{r variable-selection}
#Backward elimination
fitbmiz06full <-lm(bmiz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = malnutritionDataSet)
bestfitbmiz06<-step(fitbmiz06full, direction = "both", trace=F)
summary(bestfitbmiz06)

fithaz06full <-lm(haz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = malnutritionDataSet)
bestfithaz06<-step(fithaz06full, direction = "both", trace=F)
summary(bestfithaz06)

fitwaz06full <-lm(waz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = malnutritionDataSet)
bestfitwaz06<-step(fitwaz06full, direction = "both", trace=F)
summary(bestfitwaz06)

fitwhz06full <-lm(whz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = malnutritionDataSet)
bestfitwhz06<-step(fitwhz06full, direction = "both", trace=F) 
summary(bestfitwhz06)

```


### Prediction and Model Assessment

* Distinguish between the full model and the reduced model obtained during stepwise regression using mean error(ME), root mean square error(RMSE) and the mean absolute percentage error(MAPE)

**Model Selection**

Subsetting data into training and test dataset
```{r trained-test-data}
training_sample <- sample(1:nrow(malnutritionDataSet), size = 0.7*nrow(malnutritionDataSet),replace = F) #slect 70% sample of data for training the model

#Draw the actual training data from the dataframe
training_data<-malnutritionDataSet[training_sample,] 
training_data<-na.omit(training_data)
test_data<-malnutritionDataSet[-training_sample,] #30% is used as test dataset
test_data<-na.omit(test_data)

```

*Full and Reduced Model for Each Dependent Variable*

```{r}
bmiz06full <-lm(bmiz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = training_data)

bestfit_bmiz06_reduced<-step(bmiz06full, direction = "both", trace=F)
summary(bestfit_bmiz06_reduced)

haz06full <-lm(haz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = training_data)

bestfit_haz06_reduced<-step(haz06full, direction = "both", trace=F)
summary(bestfit_haz06_reduced)

waz06full <-lm(waz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = training_data)

bestfit_waz06_reduced<-step(waz06full, direction = "both", trace=F)
summary(bestfit_waz06_reduced)

whz06full <-lm(whz06 ~v012+v024+v025+v106+v130+v131+v151+v152+v191+bidx+bord+b4, data = training_data)

bestfit_whz06_reduced<-step(whz06full, direction = "both", trace=F) 
summary(bestfit_whz06_reduced)

```

Prediction and assessing ME, RMSE & MAPE for all dependent variables

```{r prediction-assessment-of-model}
test_data_bmiz06full<-predict(bmiz06full, data=test_data)
test_data_bmiz06reduced<-predict(bestfit_bmiz06_reduced, data=test_data)


test_data_haz06full<-predict(haz06full, data=test_data)
test_data_haz06reduced<-predict(bestfit_haz06_reduced, data=test_data)


test_data_waz06full<-predict(waz06full, data=test_data)
test_data_waz06reduced<-predict(bestfit_waz06_reduced, data=test_data)

test_data_whz06full<-predict(whz06full, data=test_data)
test_data_whz06reduced<-predict(bestfit_whz06_reduced, data=test_data)

```

**Computation of Mean Error**

```{r warning=F}
mean(test_data$bmiz06-test_data_bmiz06full)
mean(test_data$bmiz06-test_data_bmiz06reduced)

mean(test_data$haz06-test_data_haz06full)
mean(test_data$haz06-test_data_haz06reduced)

mean(test_data$waz06-test_data_waz06full)
mean(test_data$waz06-test_data_waz06reduced)

mean(test_data$whz06-test_data_whz06full)
mean(test_data$whz06-test_data_whz06reduced)


```


**Computation of RMES**
```{r warning=F}
RMSE(test_data_bmiz06full, test_data$bmiz06)
RMSE(test_data_bmiz06reduced, test_data$bmiz06)

RMSE(test_data_haz06full, test_data$haz06)
RMSE(test_data_haz06reduced, test_data$haz06)

RMSE(test_data_waz06full, test_data$waz06)
RMSE(test_data_waz06reduced, test_data$waz06)

RMSE(test_data_whz06full, test_data$whz06)
RMSE(test_data_whz06reduced, test_data$whz06)

```

**Computation of MAPE**
```{r warning=FALSE}
#bmiz06 
with(test_data %>% filter(bmiz06 > 0),
     sum(abs((bmiz06 - test_data_bmiz06full)/bmiz06))/5671 * 100 # full model
     )
with(test_data %>% filter(bmiz06 > 0),
     sum(abs((bmiz06 - test_data_bmiz06reduced)/bmiz06))/5671 * 100 # reduced model
     )

#haz06
with(test_data %>% filter(haz06 > 0),
     sum(abs((haz06 - test_data_haz06full)/haz06))/5671 * 100 # full model
     )
with(test_data %>% filter(haz06 > 0),
     sum(abs((haz06 - test_data_haz06reduced)/haz06))/5671 * 100 # reduced model
     )

#waz06
with(test_data %>% filter(waz06 > 0),
     sum(abs((waz06 - test_data_waz06full)/waz06))/5671 * 100 # full model
     )
with(test_data %>% filter(waz06 > 0),
     sum(abs((waz06 - test_data_waz06reduced)/waz06))/5671 * 100 # reduced model
     )

#whz06 
with(test_data %>% filter(whz06 > 0),
     sum(abs((whz06 - test_data_whz06full)/whz06))/5671 * 100 # full model
     )
with(test_data %>% filter(whz06 > 0),
     sum(abs((whz06 - test_data_whz06reduced)/whz06))/5671 * 100 # reduced model
     )



```




# **Results and Interpretation**

Length/height-for-age Z-score(haz06), Weight-for-age Z-score(waz06) and BMI-for-age Z-score(bmiz06) are normally distributed. From the normal plot; Weight-for-length/height Z-score is not normally distributed.

For all the response variables; atmost 12% of the variables is explained by the respective models. The model is underfitting the data.

The variability of the computed mean error(ME),RMSE and MAPE for the true and predicted values is minimal to some decimal points for response variable Length/height-for-age Z-score(haz06), Weight-for-age Z-score(waz06) and BMI-for-age Z-score(bmiz06). However, the variation of MAPE for  true and predicted values of Weight-for-length/height Z-score(whz06) is **20.6366** i.e `418.873-398.2364`

# **Conclusion**

From the analysis of respective response variables against the independent variables it is observed that only less than 12% of the outcome variables is explained by the model[Multiple R-squared are below 12%]. This is an indication that the model is underfitting the data.
The histogram and Normal probability shows that Length/height-for-age Z-score(haz06), Weight-for-age Z-score(waz06) and BMI-for-age Z-score(bmiz06) are normally distributed.

The predicted models maps the actual values accurately as demonstrated by ME, RMSE, and MAPE.

We retain the response variables Length/height-for-age Z-score(haz06), Weight-for-age Z-score(waz06) and BMI-for-age Z-score(bmiz06) because they have a smaller ME, RMSE and MAPE tests of accuracy of our model. The response variable Weight-for-length/height Z-score(whz06)is excluded as its MAPE variability is large i.e **20.6366**.



